---
title: "2. Statistical models"
author: "Sam Kane Jiménez"
output: html_document
---

```{r, setup}
knitr::opts_chunk$set(echo=FALSE,
                      warning=FALSE,
                      message=FALSE,
                      out.width='40%',
                      fig.dim=c(6,4))
```

# Statistical Models

This Rmd file covers the statistical models that I created, ran, and interpreted for this project.

Now that the data is in a neatly analyzable format, it's time to run multiple linear regression (MLR) models. These models use the *thesis* dataframe.

*Note:* Before kicking off, I've made sure to set my working directory to the folder where this Rmd file and the datasets I will be using are all saved. I then load in the R packages that I will need throughout the entire research process. I also load in the datasets created in the previous section:

```{r}
library(descr)
library(car)
library(carData)
library(stats)
library(MASS)
library(readxl)
library(tidyverse)
library(tidyr)
library(knitr)
library(ggplot2)
library(dplyr)
library(margins)
library(AER)
library(effects)
library(nnet)
library(nlme)
library(lme4)
library(sjlabelled)
library(eyelinker)
library(survey)
library(codebook)
library(ordinal)
library(sjstats)
library(Matrix)
library(boot)
library(visreg)
library(stringr)
library(lubridate)
library(reshape2)
library(stargazer)
library(wesanderson)

thesisCYG <- readRDS("thesisCYG.rds")

```

## Examining the variables

I start by looking at a summary of the variables and plots of their distributions to see if they require any transformations.

```{r}
# summary(thesisCYG)
# ggplot(thesisCYG, aes(year)) + geom_histogram() #numeric; slight left skew
# ggplot(thesisCYG, aes(econcrisis)) + geom_bar() #numeric; significant right skew
# ggplot(thesisCYG, aes(natural)) + geom_histogram() #numeric; right skew
# ggplot(thesisCYG, aes(count)) + geom_histogram() #numeric; right skew
# ggplot(thesisCYG, aes(binary)) + geom_bar() #binary; 25%-75% distribution 0-1 respectively
# ggplot(thesisCYG, aes(side_a)) + geom_bar()
# ggplot(thesisCYG, aes(side_b)) + geom_bar()
# ggplot(thesisCYG, aes(active_year)) + geom_bar() #binary; large majority = 1
# ggplot(thesisCYG, aes(type_of_violence)) + geom_bar() #ordinal; mostly 1, 2 and 3 at similar levels
# ggplot(thesisCYG, aes(population)) + geom_histogram()
# ggplot(thesisCYG, aes(logpop)) + geom_histogram()
# ggplot(thesisCYG, aes(sqkm)) + geom_histogram() #numeric; right skew
# ggplot(thesisCYG, aes(cyg_fatalities)) + geom_histogram()
# ggplot(thesisCYG, aes(cyg_civ)) + geom_histogram()
# ggplot(thesisCYG, aes(perpdeaths)) + geom_histogram()
# ggplot(thesisCYG, aes(nonperpdeaths)) + geom_histogram()
# ggplot(thesisCYG, aes(deaths_unknown)) + geom_histogram()
# ggplot(thesisCYG, aes(fat_pop)) + geom_histogram()
# ggplot(thesisCYG, aes(civcas_pop)) + geom_histogram()
# ggplot(thesisCYG, aes(fatalities_lag)) + geom_histogram()
# ggplot(thesisCYG, aes(civcas_lag)) + geom_histogram()
# ggplot(thesisCYG, aes(deaths_a)) + geom_histogram()
# ggplot(thesisCYG, aes(deaths_b)) + geom_histogram()
```

Based on those findings, I can make the appropriate transformations, plotting them out to ensure transformation improves the distribution's shape:

```{r}
thesisCYG$sqkm.log <- log(thesisCYG$sqkm)
hist(thesisCYG$sqkm)
hist(thesisCYG$sqkm.log)
ggplot(thesisCYG, aes(sqkm.log)) + geom_histogram() 

thesisCYG$econcrisis.x <- 1/(thesisCYG$econcrisis)
ggplot(thesisCYG, aes(econcrisis.x)) + geom_histogram()

thesisCYG$count.log <- log(thesisCYG$count + 0.001)
ggplot(thesisCYG, aes(count.log)) + geom_histogram() 

thesisCYG$natural.log <- log(thesisCYG$natural + 0.001)
hist(thesisCYG$natural)
ggplot(thesisCYG, aes(natural.log)) + geom_histogram() 

thesisCYG$fat_pop.log <- log(thesisCYG$fat_pop + 0.001)
ggplot(thesisCYG, aes(fat_pop.log)) + geom_histogram() 

thesisCYG$civcas_pop.log <- log(thesisCYG$civcas_pop + 0.001)
ggplot(thesisCYG, aes(civcas_pop.log)) + geom_histogram() 

thesisCYG$cyg_fatalities.log <- log(thesisCYG$cyg_fatalities + 0.001)
ggplot(thesisCYG, aes(cyg_fatalities.log)) + geom_histogram() 

thesisCYG$cyg_civ.log <- log(thesisCYG$cyg_civ + 0.001)
ggplot(thesisCYG, aes(cyg_civ.log)) + geom_histogram() 

thesisCYG$perpdeaths.log <- log(thesisCYG$perpdeaths + 0.001)
ggplot(thesisCYG, aes(perpdeaths.log)) + geom_histogram() 

thesisCYG$nonperpdeaths.log <- log(thesisCYG$nonperpdeaths + 0.001)
ggplot(thesisCYG, aes(nonperpdeaths.log)) + geom_histogram() 

thesisCYG$deaths_unknown.log <- log(thesisCYG$deaths_unknown + 0.001)
ggplot(thesisCYG, aes(deaths_unknown.log)) + geom_histogram() 
```

### Comparing each DV against each IV and among each other

To test for possible issues with multicollinearity and curvature, I compare each DV against each IV, and each DV against remaining DVs, respectively.

*DV = fat_pop.log*

Here are the comparisons for the first DV:

```{r}
# ##Categorical vars
# ggplot(thesisCYG, aes(x = active_year, y = fat_pop.log)) + geom_count()
# ggplot(thesisCYG, aes(x = binary, y = fat_pop.log)) + geom_count()
# ggplot(thesisCYG, aes(x = region, y = fat_pop.log)) + geom_count()
# ggplot(data = thesisCYG, aes(x=binary, y=fat_pop.log)) + geom_boxplot() + labs(
#     x = "Crisis Presence",
#     y = "Logged Fatality Rate")
# 
# ##Numeric vars
# ggplot(thesisCYG, aes(x = year, y = fat_pop.log)) + geom_boxplot(aes(group=fat_pop.log))
# ggplot(thesisCYG, aes(x = econcrisis.log, y = fat_pop.log)) + geom_boxplot(aes(group=fat_pop.log))
# ggplot(thesisCYG, aes(x = natural.log, y = fat_pop.log)) + geom_boxplot(aes(group=fat_pop.log))
# ggplot(thesisCYG, aes(x = count.log, y = fat_pop.log)) + geom_boxplot(aes(group=fat_pop.log))
# ggplot(thesisCYG, aes(x = civcas_pop.log, y = fat_pop.log)) + geom_boxplot(aes(group=fat_pop.log))
# ggplot(thesisCYG, aes(x = perpdeaths.log, y = fat_pop.log)) + geom_boxplot(aes(group=fat_pop.log))
# ggplot(thesisCYG, aes(x = nonperpdeaths.log, y = fat_pop.log)) + geom_boxplot(aes(group=fat_pop.log))
# ggplot(thesisCYG, aes(x = sqkm.log, y = fat_pop.log)) + geom_boxplot(aes(group=fat_pop.log))
```

*DV = civcas_pop.log*

And for the second DV:

```{r}
##Categorical vars
# ggplot(thesisCYG, aes(x = active_year, y = civcas_pop.log)) + geom_count() #
# ggplot(thesisCYG, aes(x = binary, y = civcas_pop.log)) + geom_count() #
# ggplot(thesisCYG, aes(x = region, y = civcas_pop.log)) + geom_count() #
# ggplot(data = thesisCYG, aes(x=binary, y=civcas_pop.log)) + geom_boxplot() + labs(
#     x = "Crisis Presence",
#     y = "Logged Civilian Casualty Rate")
# 
# ##Numeric vars
# ggplot(thesisCYG, aes(x = year, y = civcas_pop.log)) + geom_boxplot(aes(group=civcas_pop.log))
# ggplot(thesisCYG, aes(x = econcrisis.x, y = civcas_pop.log)) + geom_boxplot(aes(group=civcas_pop.log))
# ggplot(thesisCYG, aes(x = natural.log, y = civcas_pop.log)) + geom_boxplot(aes(group=civcas_pop.log))
# ggplot(thesisCYG, aes(x = count.log, y = civcas_pop.log)) + geom_boxplot(aes(group=civcas_pop.log))
# ggplot(thesisCYG, aes(x = fat_pop.log, y = civcas_pop.log)) + geom_boxplot(aes(group=civcas_pop.log))
# ggplot(thesisCYG, aes(x = perpdeaths.log, y = civcas_pop.log)) + geom_boxplot(aes(group=civcas_pop.log))
# ggplot(thesisCYG, aes(x = nonperpdeaths.log, y = civcas_pop.log)) + geom_boxplot(aes(group=civcas_pop.log))
# ggplot(thesisCYG, aes(x = sqkm.log, y = civcas_pop.log)) + geom_boxplot(aes(group=civcas_pop.log))
```

*T-tests and F-tests for the impact of each IV on the DV*

```{r}
attach(thesisCYG)
t.test(year, fat_pop.log)
t.test(natural.log, fat_pop.log)
t.test(count.log, fat_pop.log)
t.test(civcas_pop.log, fat_pop.log)
t.test(perpdeaths.log, fat_pop.log)
t.test(nonperpdeaths.log, fat_pop.log)
t.test(sqkm.log, fat_pop.log)
```

## Multiple linear regression models

I run several models, with and without lags and with and without regional controls.

*Models without lags with regional controls*

```{r}
binary.fat <- lm(fat_pop.log ~ year + sqkm.log + region + active_year + binary) #impact of crisis vs. not on total fatalities
summary(binary.fat)
mbinary.fat <- margins(binary.fat)
plot(mbinary.fat)

binary.civ <- lm(civcas_pop.log ~ year + sqkm.log + region + active_year + binary) #impact of crisis vs. not on civilian fatalities
summary(binary.civ)
count.fat <- lm(fat_pop.log ~ year + sqkm.log + region + active_year + count) #impact of crisis tally on total fatalities
summary(count.fat)
count.civ <- lm(civcas_pop.log ~ year + sqkm.log + region + active_year + count) #impact of crisis tally on civilian fatalities
summary(count.civ)
econ.fat <- lm(fat_pop.log ~ year + sqkm.log + region + active_year + econcrisis) #impact of economic crises on total fatalities
summary(econ.fat)
econ.civ <- lm(civcas_pop.log ~ year + sqkm.log + region + active_year + econcrisis) #impact of economic crises on total fatalities
summary(econ.civ)
nat.fat <- lm(fat_pop.log ~ year + sqkm.log + region + active_year + natural) #impact of natural crises on total fatalities
summary(nat.fat)
nat.civ <- lm(civcas_pop.log ~  year + sqkm.log + region + active_year + natural) #impact of natural crises on total fatalities
summary(nat.civ)
```

*Models without lags or regional controls*

```{r}
binary.fat.c <- lm(fat_pop.log ~  year + sqkm.log + active_year + binary) #impact of crisis vs. not on total fatalities
summary(binary.fat.c)
binary.civ.c <- lm(civcas_pop.log ~ year + sqkm.log + active_year + binary) #impact of crisis vs. not on civilian fatalities
summary(binary.civ.c)
count.fat.c <- lm(fat_pop.log ~ year + sqkm.log + active_year + count) #impact of crisis tally on total fatalities
summary(count.fat.c)
count.civ.c <- lm(civcas_pop.log ~ year + sqkm.log + active_year + count) #impact of crisis tally on civilian fatalities
summary(count.civ.c)
econ.fat.c <- lm(fat_pop.log ~ year + sqkm.log + active_year + econcrisis) #impact of economic crises on total fatalities
summary(econ.fat.c)
econ.civ.c <- lm(civcas_pop.log ~ year + sqkm.log + active_year + econcrisis) #impact of economic crises on total fatalities
summary(econ.civ.c)
nat.fat.c <- lm(fat_pop.log ~ year + sqkm.log + active_year + natural) #impact of natural crises on total fatalities
summary(nat.fat.c)
nat.civ.c <- lm(civcas_pop.log ~ year + sqkm.log + active_year + natural) #impact of natural crises on total fatalities
summary(nat.civ.c)
```

*Models with lags and regional controls*

```{r}
binary.fatlag <- lm(fatalities_lag ~ binary + year + region + active_year + sqkm.log) #impact of crisis vs. not on total fatalities
summary(binary.fatlag)
binary.civlag <- lm(civcas_lag ~ binary + year + region + active_year + sqkm.log) #impact of crisis vs. not on civilian fatalities
summary(binary.civlag)
count.fatlag <- lm(fatalities_lag ~ count + year + region + active_year + sqkm.log) #impact of crisis tally on total fatalities
summary(count.fatlag)
count.civlag <- lm(civcas_lag ~ count + year + region + active_year + sqkm.log) #impact of crisis tally on civilian fatalities
summary(count.civlag)
econ.fatlag <- lm(fatalities_lag ~ econcrisis + year + active_year + region + sqkm.log) #impact of economic crises on total fatalities
summary(econ.fatlag)
econ.civlag <- lm(civcas_lag ~ econcrisis + year + region + active_year + sqkm.log) #impact of economic crises on total fatalities
summary(econ.civlag)
nat.fatlag <- lm(fatalities_lag ~ natural + year + region + active_year + sqkm.log) #impact of natural crises on total fatalities
summary(nat.fatlag)
nat.civlag <- lm(civcas_lag ~ natural + year + region + active_year + sqkm.log) #impact of natural crises on total fatalities
summary(nat.civlag)
```

*Models with lags without regional controls*

```{r}
binary.fatlag.c <- lm(fatalities_lag ~ binary + year + sqkm.log + active_year) #impact of crisis vs. not on total fatalities
summary(binary.fatlag.c)
binary.civlag.c <- lm(civcas_lag ~ binary + year + sqkm.log + active_year) #impact of crisis vs. not on civilian fatalities
summary(binary.civlag.c)
count.fatlag.c <- lm(fatalities_lag ~ count + year + sqkm.log + active_year) #impact of crisis tally on total fatalities
summary(count.fatlag.c)
count.civlag.c <- lm(civcas_lag ~ count + year + sqkm.log + active_year) #impact of crisis tally on civilian fatalities
summary(count.civlag.c)
econ.fatlag.c <- lm(fatalities_lag ~ econcrisis + year + sqkm.log + active_year) #impact of economic crises on total fatalities
summary(econ.fatlag.c)
econ.civlag.c <- lm(civcas_lag ~ econcrisis + year + sqkm.log + active_year) #impact of economic crises on total fatalities
summary(econ.civlag.c)
nat.fatlag.c <- lm(fatalities_lag ~ natural + year + sqkm.log + active_year) #impact of natural crises on total fatalities
summary(nat.fatlag.c)
nat.civlag.c <- lm(civcas_lag ~ natural + year + sqkm.log + active_year) #impact of natural crises on total fatalities
summary(nat.civlag.c)

natecon.fat <- lm(fat_pop.log ~ year + sqkm.log + region + active_year + natural + econcrisis) #impact of natural crises on total fatalities
summary(natecon.fat)
natecon.civ <- lm(civcas_pop.log ~  year + sqkm.log + region + active_year + natural + econcrisis) #impact of natural crises on total fatalities
summary(natecon.civ)
natecon.fat.c <- lm(fat_pop.log ~ year + sqkm.log + active_year + natural + econcrisis) #impact of natural crises on total fatalities
summary(natecon.fat.c)
natecon.civ.c <- lm(civcas_pop.log ~ year + sqkm.log + active_year + natural + econcrisis) #impact of natural crises on total fatalities
summary(natecon.civ.c)
natecon.fatlag <- lm(fatalities_lag ~ natural + econcrisis + year + region + active_year + sqkm.log) #impact of natural crises on total fatalities
summary(natecon.fatlag)
natecon.civlag <- lm(civcas_lag ~ natural + econcrisis + year + region + active_year + sqkm.log) #impact of natural crises on total fatalities
summary(natecon.civlag)
natecon.fatlag.c <- lm(fatalities_lag ~ natural +econcrisis + year + sqkm.log + active_year) #impact of natural crises on total fatalities
summary(natecon.fatlag.c)
natecon.civlag.c <- lm(civcas_lag ~ natural + econcrisis + year + sqkm.log + active_year) #impact of natural crises on total fatalities
summary(natecon.civlag.c)

detach(thesisCYG)
```

## Export regression tables

The following code exports the regression tables in HTML format.

```{r}
stargazer(binary.fat, binary.fat.c, count.fat, count.fat.c, econ.fat, #econ.fat.c, nat.fat, nat.fat.c,
          type="html",
          align=TRUE,
          report="vc*",
          title="Logged Fatality Rates as a Function of Crises",
          dep.var.caption  = "Dependent Variable: Logged Fatality Rate",
          dep.var.labels="Model Focus",
          column.labels   = c("Crisis Presence", "Crisis Count", "Economic Crisis", "Natural Disaster"),
          column.separate = c(2, 2, 2, 2),
          covariate.labels=c("Year","Logged Territory","America", "Asia", "Europe", "Middle East", "Active Conflict", "Crisis Presence", "Crisis Count", "Economic Crisis", "Natural Disaster"),
          out="MLR_fatalities_outputcyg.htm")

stargazer(binary.civ, binary.civ.c, count.civ, count.civ.c, econ.civ, #econ.civ.c, nat.civ, nat.civ.c,
          type="html",
          align=TRUE,
          report="vc*",
          title="Logged Civilian Casualty Rates as a Function of Crises",
          dep.var.caption  = "Dependent Variable: Logged Civilian Casualty Rate",
          dep.var.labels="Model Focus",
          column.labels   = c("Crisis Presence", "Crisis Count", "Economic Crisis", "Natural Disaster"),
          column.separate = c(2, 2, 2, 2),
          covariate.labels=c("Year","Logged Territory","America", "Asia", "Europe", "Middle East", "Active Conflict", "Crisis Presence", "Crisis Count", "Economic Crisis", "Natural Disaster"),
          out="MLR_casualties_outputcyg.htm")

stargazer(binary.fatlag, binary.fatlag.c, count.fatlag, count.fatlag.c, #econ.fatlag, econ.fatlag.c, nat.fatlag, nat.fatlag.c,
          type="html",
          align=TRUE,
          report="vc*",
          title="Change in Fatality Rate as a Function of Crises",
          dep.var.caption  = "Dependent Variable: 1-Year Change in Fatality Rate",
          dep.var.labels="Model Focus",
          column.labels   = c("Crisis Presence", "Crisis Count", "Economic Crisis", "Natural Disaster"),
          column.separate = c(2, 2, 2, 2),
          covariate.labels=c("Crisis Presence","Crisis Count","Economic Crisis", "Natural Disaster", "Year", "America", "Asia", "Europe", "Middle East", "Active Conflict", "Logged Territory"),
          out="MLR_fatalities_laggedcyg.htm")

stargazer(binary.civlag, binary.civlag.c, count.civlag, count.civlag.c, #econ.civlag, econ.civlag.c, nat.civlag, nat.civlag.c,
          type="html",
          align=TRUE,
          report="vc*",
          title="Change in Civilian Casualty Rate as a Function of Crises",
          dep.var.caption  = "Dependent Variable: 1-Year Change in Civilian Casualty Rate",
          dep.var.labels="Model Focus",
          column.labels   = c("Crisis Presence", "Crisis Count", "Economic Crisis", "Natural Disaster"),
          column.separate = c(2, 2, 2, 2),
          covariate.labels=c("Crisis Presence","Crisis Count","Economic Crisis", "Natural Disaster", "Year", "America", "Asia", "Europe", "Middle East", "Active Conflict", "Logged Territory"),
           out="MLR_casualties_laggedcyg.htm")

```
