---
title: "4. Case study selection"
author: "Sam Kane Jiménez"
output: html_document
---

```{r, setup}
knitr::opts_chunk$set(echo=FALSE,
                      warning=FALSE,
                      message=FALSE,
                      out.width='40%',
                      fig.dim=c(6,4))
```

# Case study selection and plots

For this project's qualitative component, I randomly selected three cases to analyze, stratified by geographic region. This section covers the selection process as well as the plots and figures I created to support the discussion of each case study.

*Note:* Before kicking off, I've made sure to set my working directory to the folder where this Rmd file and the datasets I will be using are all saved. I then load in the R packages that I will need throughout the entire research process. I also load in the datasets created in the previous section:

*Note:* Before kicking off, I've made sure to set my working directory to the folder where this Rmd file and the datasets I will be using are all saved. I then load in the R packages that I will need throughout the entire research process. I also load in the datasets created in the previous section:

```{r}
library(descr)
library(car)
library(carData)
library(stats)
library(MASS)
library(readxl)
library(tidyverse)
library(tidyr)
library(knitr)
library(ggplot2)
library(dplyr)
library(margins)
library(AER)
library(effects)
library(nnet)
library(nlme)
library(lme4)
library(sjlabelled)
library(eyelinker)
library(survey)
library(codebook)
library(ordinal)
library(sjstats)
library(Matrix)
library(boot)
library(visreg)
library(stringr)
library(lubridate)
library(reshape2)
library(stargazer)
library(wesanderson)

emdat <- readRDS("emdat.rds")
thesisCYG <- readRDS("thesisCYG.rds")

```

## Case selection

To select my qualitative cases to study, I subset *emdat* to natural disaster cases that affected more than 30,000 people and that occurred after 1988. I grouped the dataset's regional subsets into my own regions of interest - Africa, America, Asia, and the Middle East and Northern Africa - and dropped the remaining categories.

```{r}
emdatopts <- subset(emdat, emdat$TotalAffected >= 30000)
emdatopts$Year <- as.numeric(emdatopts$year)
emdatopts <- subset(emdatopts, emdatopts$Year > 1988)
```

At the time, I used an online random number generator to select the corresponding case in each regional subset, using the following code:

```{r}
# asia <- subset(emdatopts, emdatopts$Region == "Central Asia" | emdatopts$Region == "Melanesia" | emdatopts$Region == "Southern Asia" | emdatopts$Region == "South-Eastern Asia")
# africa <- subset(emdatopts, emdatopts$Region == "Southern Africa" | emdatopts$Region == "Western Africa" | emdatopts$Region == "Middle Africa" | emdatopts$Region == "Eastern Africa")
# america <- subset(emdatopts, emdatopts$Region == "South America" | emdatopts$Region == "Northern America" | emdatopts$Region == "Central America")
# mena <- subset(emdatopts, emdatopts$Region == "Western Asia" | emdatopts$Region == "Northern Africa")
```

Upon reviewing this document, I realized I could employ R to conduct a stratified random sampling of the data as follows:

```{r}
#Recategorize the regions per my preferences
emdatopts$Region[emdatopts$Region == "Central Asia" | emdatopts$Region == "Melanesia" | emdatopts$Region == "Southern Asia" | emdatopts$Region == "South-Eastern Asia" | emdatopts$Region == "Eastern Asia"] <- "Asia"

emdatopts$Region[emdatopts$Region == "Southern Africa" | emdatopts$Region == "Western Africa" | emdatopts$Region == "Middle Africa" | emdatopts$Region == "Eastern Africa"] <- "Africa"

emdatopts$Region[emdatopts$Region == "South America" | emdatopts$Region == "Northern America" | emdatopts$Region == "Central America" | emdatopts$Region == "Caribbean"] <- "America"

emdatopts$Region[emdatopts$Region == "Western Asia" | emdatopts$Region == "Northern Africa"] <- "MENA"

#drop other regions
emdatopts2 <- emdatopts %>% 
  subset(Region %in% c("Asia", "Africa", "America", "MENA"))

#randomly select one case from each regional category
casestudies <- emdatopts %>%
  group_by(Region) %>%
  sample_n(size=1)

#randomly select 3 case studies from the 4, owing to my time constraints for research and writing
casestudies2 <- casestudies[sample(nrow(casestudies), size=3), ]
```

Regardless of the selection method, there was some trial and error, as I had to corroborate that the selected cases overlapped with places where ANSAs were active at the time.

After selecting my cases, I created plots to illustrate ANSA violence levels before and after each disaster.

For each case study, I subset cases in *thesisCYG* to only include those involving the pertinent ANSA and its country of operations to then make an attack frequency chart with time (start dates) on the X axis and casualties on the Y axis:

## Gulf Cartel and Huracán Alex, Mexico 2010

```{r}
gulf <- subset(thesisCYG, str_detect(thesisCYG$cyg, "Gulf Cartel"))
gulf$year <- as.factor(gulf$year)

breakV <- seq(2004,2019,by=2) #start and end years for the chart, counted by 2s
gulf2 <- gulf %>% 
  dplyr::select(year, deaths_a, deaths_b, cyg_civ, deaths_unknown) #select pertinent variables for plot
gulf2$cyg <- NULL
meltdf2 <- melt(gulf2,id="year")

#create plot
ggplot(meltdf2,aes(fill=variable, x=year,y=value,group=variable)) + geom_bar(stat = "identity", position = "stack") +  scale_fill_manual(name = "Group", labels = c("Civilians", "Government Forces", "Gulf Cartel Forces", "Unknown"), values = (wes_palette("GrandBudapest1"))) + scale_x_discrete(name ="Year", breaks = breakV) + scale_y_continuous(name ="Number of Fatalities") + theme(title = element_text(size=9), axis.text.x = element_text(size=6), axis.text.y = element_text(size=6), axis.title.x = element_text(size=7), axis.title.y = element_text(size=7), legend.title = element_text(size = 7), legend.key.size = unit(.5,"line"), legend.text = element_text(size=6), legend.position = "bottom") + labs(y = "Number of Deaths", title = "Deaths by Group in Gulf Cartel-Related Attacks, 2004-2019") + annotate("rect", xmin=7, xmax=7.5, ymin=-Inf, ymax=Inf, alpha=0.2, fill="red")
```

## PKK and the Van Earthquake, Turkey 1998

```{r}
#make an attack frequency chart with time (start dates) on the X axis and casualties on the Y axis
#subset to only cases including "PKK" in dyad_name
pkk <- subset(thesisCYG, str_detect(thesisCYG$cyg, "PKK"))
pkk$year <- as.factor(pkk$year)

breakV <- seq(1989,2019,by=4) #window of years to show on chart, separated by 4s
pkk2 <- pkk %>% 
  dplyr::select(year, deaths_a, deaths_b, cyg_civ, deaths_unknown) #select pertinent variables
pkk2$cyg <- NULL
meltdf2 <- melt(pkk2,id="year")

#create plot
ggplot(meltdf2,aes(fill=variable, x=year,y=value,group=variable)) + geom_bar(stat = "identity", position = "stack") +  scale_fill_manual(name = "Group", labels = c("Civilians", "Government Forces", "PKK Forces", "Unknown"), values = (wes_palette("GrandBudapest1"))) + scale_x_discrete(name ="Year", breaks = breakV) + scale_y_continuous(name ="Number of Fatalities") + theme(title = element_text(size=9), axis.text.x = element_text(size=6), axis.text.y = element_text(size=6), axis.title.x = element_text(size=7), axis.title.y = element_text(size=7), legend.title = element_text(size = 7), legend.key.size = unit(.5,"line"), legend.text = element_text(size=6), legend.position = "bottom") + labs(y = "Number of Deaths", title = "Deaths by Group in PKK-Related Attacks, 1989-2019") + annotate("rect", xmin=23, xmax=23.5, ymin=-Inf, ymax=Inf, alpha=0.2, fill="red")
```

## ONLF and the drought, Ethiopia 2009

```{r}
onlf <- subset(thesisCYG, perpetrator=="ONLF" & country=="Ethiopia") #subset to ANSA attack information relevant to case study
onlf <- onlf %>% ungroup() %>% add_row(tibble_row(year=1997, active_year=0))
onlf$year <- as.factor(onlf$year)

onlf$deaths_a[onlf$ISOyear == "ETH 2003" & onlf$type_of_violence==3] <- 0
onlf$cyg_civ[onlf$ISOyear == "ETH 2003" & onlf$type_of_violence==3] <- 2 
onlf$deaths_b[onlf$ISOyear == "ETH 2003" & onlf$type_of_violence==3] <- 0
onlf$deaths_a[onlf$ISOyear == "ETH 2007" & onlf$type_of_violence==3] <- 0
onlf$cyg_civ[onlf$ISOyear == "ETH 2007" & onlf$type_of_violence==3] <- 82 
onlf$deaths_b[onlf$ISOyear == "ETH 2007" & onlf$type_of_violence==3] <- 0

breakV <- seq(1988,2018,by=2) #window of years to show on chart, separated by 2s
onlf2 <- onlf %>% 
  dplyr::select(year, deaths_a, deaths_b, cyg_civ, deaths_unknown) #keep pertinent variables
onlf2$cyg <- NULL
meltdf2 <- melt(onlf2,id="year")

#create plot
ggplot(meltdf2,aes(fill=variable, x=year,y=value,group=variable)) + geom_bar(stat = "identity", position = "stack") +  scale_fill_manual(name = "Group", labels = c("Civilians", "Government Forces", "ONLF Forces", "Unknown"), values = (wes_palette("GrandBudapest1"))) + scale_x_discrete(name ="Year", breaks = breakV) + scale_y_continuous(name ="Number of Fatalities") + theme(title = element_text(size=9), axis.text.x = element_text(size=6), axis.text.y = element_text(size=6), axis.title.x = element_text(size=7), axis.title.y = element_text(size=7), legend.title = element_text(size = 7), legend.key.size = unit(.5,"line"), legend.text = element_text(size=6), legend.position = "bottom") + labs(y = "Number of Deaths", title = "Deaths by Group in ONLF-Related Attacks, 1994-2018") + annotate("rect", xmin=14.3, xmax=18.6, ymin=-Inf, ymax=Inf, alpha=0.2, fill="red")
```
